Estimating process-based model parameters from species distribution data using the evolutionary algorithm CMA-ES

### 2023-04-18 manuscript/paper.Rmd
removed green annotations for reviewers

---
output: 
  stevetemplates::article:
    keep_tex: true
    fig_caption: true
    number_sections: yes
bibliography: bib/paper.bib
csl: bib/mee.csl
title: "Estimating process-based model parameters from species distribution data using the evolutionary algorithm CMA-ES"
author:
- name: Victor Van der Meersch \footnote{Corresponding author - victor.vandermeersch@cefe.cnrs.fr}, Isabelle Chuine
  affiliation: CEFE, Université de Montpellier, CNRS, EPHE, IRD, Montpellier, France
abstract: "1. Two main types of species distribution models are used to project species range shifts in future climatic conditions: correlative and process-based models. Although there is some continuity between these two types of models, they are fundamentally different in their hypotheses (statistical relationships *vs* mechanistic relationships) and their calibration methods \\textcolor{darkgreen}{(SDMs tend to be occurrence data-driven while PBMs tend to be prior-driven)}. \n\n 2. One of the limitations to the use of process-based models is the difficulty to parameterize them for a large number of species \\textcolor{darkgreen}{compared to correlative SDMs}. We investigated the feasibility of using an evolutionary algorithm (called covariance matrix adaptation evolution strategy, CMA-ES) to calibrate process-based models using species distribution data. This method is well established in some fields (robotics, aerospace research, ...), but has never been used, to our knowledge, in ecology, despite its ability to deal with very large space dimensions. Using tree species occurrence data across Europe, we adapted the CMA-ES algorithm to find appropriate values of model parameters. We estimated simultaneously 27 to 77 parameters of two process-based models simulating forest tree's ecophysiology for three species with varying range sizes and geographical distributions. \n\n 3. CMA-ES provided parameter estimates leading to better prediction of species distribution than parameter estimates based on experts knowledge. Our results also revealed that some model parameters and processes were strongly dependent, and different parameter combinations could therefore lead to high model accuracy. \n\n 4. \\textcolor{darkgreen}{We conclude that} CMA-ES is an efficient state-of-the-art method to calibrate process-based models with a large number of parameters using species occurrence data. Inverse modelling using CMA-ES is a powerful method to calibrate process-based parameters which can hardly be measured. However, the method does not warranty that parameter estimates are correct because of several sources of bias, similarly to correlative models, and expert knowledge is required to validate results. \n\n"
keywords: "calibration, evolutionary algorithm, cma-es, species distribution model, process-based model, trees"
date: "`r format(Sys.time(), '%B %d, %Y')`"
geometry: margin=1in
fontfamily: libertine
fontsize: 11pt
spacing: double
endnote: no
biblatex: true
link-citations: true

@@

Calibration (i.e. parameter setting and estimation) is a fundamental step in the modelling process. It allows the model to reproduce the reality with more or less success. The result of the calibration provides insights on the ability of the model to reproduce and explain reality (model predictive power). Calibration of complex models such as process-based SDMs is time-consuming, and modelers are often challenged by the dimension of the parameter space, the complexity of the possible correlations among parameters, and the scarcity of observed experimental data to calibrate them. Therefore, although the philosophy of such models is to measure parameters, statistical inference might be useful when data are not yet available to infer parameter values. Parameter inference can be achieved through several methods which have been developed in the last decades. \textcolor{darkgreen}{Most of them fall into two categories: informal and statistical calibration. On the one hand, statistical calibration assumes a data-generating model, and a likelihood function, which quantifies the probability of the observed data given the model parameters, is expressed mathematically. This likelihood is the foundation of \emph{maximum likelihood estimation} and \emph{Bayesian inference}, two commonly used methods for parameter estimation. In practice, deriving the likelihood function can be quite challenging for complex models with many parameters and thresholding. One approach is to use numerical methods such as \emph{simulation-based inference} or \emph{approximate Bayesian computation} [e.g. @Hartig2014]\textcolor{darkgreen}{. However, obtaining a reliable estimate of the likelihood of a complex process-based model can require a large number of simulated samples, which can be computationally expensive. On the other hand, informal calibration uses an informal objective function to measure the discrepancy between the model predictions and the observed data. By minimizing the objective function, one is able to identify the parameter values that best fit the observed data, even though it may not have the same statistical rigor as statistical calibration. This optimization step can be achieved with several methods, either deterministic (e.g. Nelder-Mead method) or stochastic (e.g. simulated annealing, evolutionary algorithms).}     
Recently, an approach belonging to the evolutionary algorithm family, called Covariance Matrix Adaptation Evolution Strategy (CMA-ES), has been proposed [@Hansen2001]. One of the advantages of CMA-ES is its ability to cumulate information over iterations in order to adapt its own parameters (in particular the covariance matrix), which makes it more robust to noise. CMA-ES is especially performant for non-separable problems (i.e. when the model parameters are dependent) and large search space. This method has been successfully applied in various fields such as aerospace [e.g. @Collange2010], optics [e.g. @Gagne2008], and robotics [e.g. @Hill2020]. CMA-ES is acknowledged to be one of the most efficient approaches in continuous black-box optimization [@Hansen2010] but to our knowledge has never been used in ecology.

@@

\textcolor{darkgreen}{Our results showed that CMA-ES is an efficient optimizer for inverse calibration of complex ecological models. The algorithm was able to find parameter sets that significantly improved the predictions of the calibrated models compared to the expert parametrization.  However, our study also highlighted the issue of non-identifiability of parameter values due to the data limitation and the dependence between process-based model parameters, which may result in diverging parameter values even though the calibrated models describe the observed species distribution well.}

@@

Here we demonstrated that inverse calibration with CMA-ES is feasible and provide good results for complex and runtime-expensive ecological models.  
With a subsampling of species occurrence data, the algorithm succeeds in finding parameter sets which provide \textcolor{darkgreen}{higher} AUC values. The predictions of the calibrated models are sharply improved compared to expert parametrization (\hyperref[fig:phenofitmaps]{Figure 1} and \hyperref[fig:castaneamaps]{Figure 2}). Two striking examples are the increase in the performance of PHENOFIT model for silver fir, from 0.72 to 0.9, and of CASTANEA model for holm oak, from 0.7 to 0.95. Moreover, CMA-ES performed equally well regardless of the species occurrence subset used during calibration (\hyperref[fig:cmaesrepAUCcal]{Figure 3.B.}), and thus permitted to find a good compromise between computational cost and calibration efficiency.

CMA-ES is a "generic" optimizer which can be applied to various problems. It is easy to use as it does not require an extensive tuning to  efficiently explore the parameter space. We only had to choose the number of candidate solutions $\lambda$, and the initial search region (initial starting point and step size $\sigma$). As well as being quasi parameter-free, CMA-ES has several structural advantages particularly useful for complex optimization problems. \textcolor{darkgreen}{First, the algorithm's covariance matrix enables the learning of second-order information, which provides insights into pairwise dependencies between parameters. Second, the covariance matrix adaptation of CMA-ES is highly efficient in handling ill-conditioned and non-separable problems }[@Hansen2011]\textcolor{darkgreen}{. Last, CMA-ES's update mechanism of step size $\sigma$ (i.e. the mutation force) helps prevent premature convergence }[@Hansen2001]\textcolor{darkgreen}{, allowing the algorithm to explore more of the search space.}
CMA-ES has been shown to outperform several other optimization algorithms [@Hansen2010], and is usually the most efficient method when the target cost (i.e. the number of objective function evaluations) is about $100*N$ [$N$ being the dimension of the parameter search space, @Baeck2013].

@@

\textcolor{darkgreen}{Equifinality is a common problem encountered during model calibration, where multiple parameter sets can produce equally good fits to the observed data. This issue can arise due to several factors, including limitations in the quantity or quality of available data, competing processes within the model, and parameter interactions that affect the model output in complex ways.}

@@

Consequently, several parameter sets may be statistically equivalent and parameters non-identifiable. In fact, calibration repetitions gave diverging parameter values (\hyperref[fig:unfoldingplots]{Figure 4}) while being efficient in distinguishing between species presence and absence (i.e. AUC ~ 0.9, \hyperref[fig:cmaesrepAUCcal]{Figure 3}). Thus, even if the calibrated model describes the observed species distribution very well, it does not necessarily mean that parameter values are ecologically relevant. This concern is similar to the criticisms against correlative SDMs, in which parameter values and correlations that well reproduce species ranges do not necessarily describe a complex biological reality. In our case, the constraints imposed by the explicit mathematical equations embedded in the models we used were not sufficient to ensure calibration convergence towards similar solutions that would have suggested a high biological realism. However, it is worth noting that we deliberately chose large parameter ranges (although biologically realistic, i.e. corresponding to the observations made on the different processes modelled across different species) in order to give free rein to the optimization algorithm. As our goal was to assess the performance of CMA-ES objectively, we did not attempt to minimize this non-identifiability issue by restricting the parameter space. To deal with equifinality issues, an avenue to explore could be the use of multiple objective functions during model calibration to assess different aspects of the model performance. \textcolor{darkgreen}{Additionally, if a closed-form likelihood can be derived, one could use a Bayesian framework to combine prior knowledge and inverse estimation of parameters to constrain the parameter space and study the nature of trade-offs between parameters} [@Hartig2012; @Cailleret2020].

@@

Our goal here was to investigate the performance of CMA-ES to calibrate quite complex process-based species distribution models using species occurrence data. We did not attempt to validate our parametrizations using temporally or spatially independent data, and AUC was only used to determine if model outputs were consistent with species distributions. \textcolor{darkgreen}{However, AUC is scale-invariant: it measures how well predictions are ranked rather than their absolute values. For example, with PHENOFIT, a species with a fitness of $0.8$ could be considered as absent while another one with the same fitness could be considered as present. Therefore, when it is used as an objective function for model calibration, we probably lack some precision and consistency among species’ parameters estimation.} Further work could thus be conducted to examine the effects of choosing a different objective function.

>>>>

---
output: 
  stevetemplates::article:
    keep_tex: true
    fig_caption: true
    number_sections: yes
bibliography: bib/paper.bib
csl: bib/mee.csl
title: "Estimating process-based model parameters from species distribution data using the evolutionary algorithm CMA-ES"
author:
- name: Victor Van der Meersch \footnote{Corresponding author - victor.vandermeersch@cefe.cnrs.fr}, Isabelle Chuine
  affiliation: CEFE, Université de Montpellier, CNRS, EPHE, IRD, Montpellier, France
abstract: "1. Two main types of species distribution models are used to project species range shifts in future climatic conditions: correlative and process-based models. Although there is some continuity between these two types of models, they are fundamentally different in their hypotheses (statistical relationships *vs* mechanistic relationships) and their calibration methods (SDMs tend to be occurrence data-driven while PBMs tend to be prior-driven). \n\n 2. One of the limitations to the use of process-based models is the difficulty to parameterize them for a large number of species compared to correlative SDMs. We investigated the feasibility of using an evolutionary algorithm (called covariance matrix adaptation evolution strategy, CMA-ES) to calibrate process-based models using species distribution data. This method is well established in some fields (robotics, aerospace research, ...), but has never been used, to our knowledge, in ecology, despite its ability to deal with very large space dimensions. Using tree species occurrence data across Europe, we adapted the CMA-ES algorithm to find appropriate values of model parameters. We estimated simultaneously 27 to 77 parameters of two process-based models simulating forest tree's ecophysiology for three species with varying range sizes and geographical distributions. \n\n 3. CMA-ES provided parameter estimates leading to better prediction of species distribution than parameter estimates based on experts knowledge. Our results also revealed that some model parameters and processes were strongly dependent, and different parameter combinations could therefore lead to high model accuracy. \n\n 4. We conclude that CMA-ES is an efficient state-of-the-art method to calibrate process-based models with a large number of parameters using species occurrence data. Inverse modelling using CMA-ES is a powerful method to calibrate process-based parameters which can hardly be measured. However, the method does not warranty that parameter estimates are correct because of several sources of bias, similarly to correlative models, and expert knowledge is required to validate results. \n\n"
keywords: "calibration, evolutionary algorithm, cma-es, species distribution model, process-based model, trees"
date: "`r format(Sys.time(), '%B %d, %Y')`"
geometry: margin=1in
fontfamily: libertine
fontsize: 11pt
spacing: double
endnote: no
biblatex: true
link-citations: true

@@

Calibration (i.e. parameter setting and estimation) is a fundamental step in the modelling process. It allows the model to reproduce the reality with more or less success. The result of the calibration provides insights on the ability of the model to reproduce and explain reality (model predictive power). Calibration of complex models such as process-based SDMs is time-consuming, and modelers are often challenged by the dimension of the parameter space, the complexity of the possible correlations among parameters, and the scarcity of observed experimental data to calibrate them. Therefore, although the philosophy of such models is to measure parameters, statistical inference might be useful when data are not yet available to infer parameter values. Parameter inference can be achieved through several methods which have been developed in the last decades. Most of them fall into two categories: informal and statistical calibration. On the one hand, statistical calibration assumes a data-generating model, and a likelihood function, which quantifies the probability of the observed data given the model parameters, is expressed mathematically. This likelihood is the foundation of \emph{maximum likelihood estimation} and \emph{Bayesian inference}, two commonly used methods for parameter estimation. In practice, deriving the likelihood function can be quite challenging for complex models with many parameters and thresholding. One approach is to use numerical methods such as \emph{simulation-based inference} or \emph{approximate Bayesian computation [e.g. @Hartig2014]. However, obtaining a reliable estimate of the likelihood of a complex process-based model can require a large number of simulated samples, which can be computationally expensive. On the other hand, informal calibration uses an informal objective function to measure the discrepancy between the model predictions and the observed data. By minimizing the objective function, one is able to identify the parameter values that best fit the observed data, even though it may not have the same statistical rigor as statistical calibration. This optimization step can be achieved with several methods, either deterministic (e.g. Nelder-Mead method) or stochastic (e.g. simulated annealing, evolutionary algorithms).     
Recently, an approach belonging to the evolutionary algorithm family, called Covariance Matrix Adaptation Evolution Strategy (CMA-ES), has been proposed [@Hansen2001]. One of the advantages of CMA-ES is its ability to cumulate information over iterations in order to adapt its own parameters (in particular the covariance matrix), which makes it more robust to noise. CMA-ES is especially performant for non-separable problems (i.e. when the model parameters are dependent) and large search space. This method has been successfully applied in various fields such as aerospace [e.g. @Collange2010], optics [e.g. @Gagne2008], and robotics [e.g. @Hill2020]. CMA-ES is acknowledged to be one of the most efficient approaches in continuous black-box optimization [@Hansen2010] but to our knowledge has never been used in ecology.

@@

Our results showed that CMA-ES is an efficient optimizer for inverse calibration of complex ecological models. The algorithm was able to find parameter sets that significantly improved the predictions of the calibrated models compared to the expert parametrization.  However, our study also highlighted the issue of non-identifiability of parameter values due to the data limitation and the dependence between process-based model parameters, which may result in diverging parameter values even though the calibrated models describe the observed species distribution well.

@@

Here we demonstrated that inverse calibration with CMA-ES is feasible and provide good results for complex and runtime-expensive ecological models.  
With a subsampling of species occurrence data, the algorithm succeeds in finding parameter sets which provide higher AUC values. The predictions of the calibrated models are sharply improved compared to expert parametrization (\hyperref[fig:phenofitmaps]{Figure 1} and \hyperref[fig:castaneamaps]{Figure 2}). Two striking examples are the increase in the performance of PHENOFIT model for silver fir, from 0.72 to 0.9, and of CASTANEA model for holm oak, from 0.7 to 0.95. Moreover, CMA-ES performed equally well regardless of the species occurrence subset used during calibration (\hyperref[fig:cmaesrepAUCcal]{Figure 3.B.}), and thus permitted to find a good compromise between computational cost and calibration efficiency.

CMA-ES is a "generic" optimizer which can be applied to various problems. It is easy to use as it does not require an extensive tuning to  efficiently explore the parameter space. We only had to choose the number of candidate solutions $\lambda$, and the initial search region (initial starting point and step size $\sigma$). As well as being quasi parameter-free, CMA-ES has several structural advantages particularly useful for complex optimization problems. First, the algorithm's covariance matrix enables the learning of second-order information, which provides insights into pairwise dependencies between parameters. Second, the covariance matrix adaptation of CMA-ES is highly efficient in handling ill-conditioned and non-separable problems [@Hansen2011]. Last, CMA-ES's update mechanism of step size $\sigma$ (i.e. the mutation force) helps prevent premature convergence [@Hansen2001], allowing the algorithm to explore more of the search space.
CMA-ES has been shown to outperform several other optimization algorithms [@Hansen2010], and is usually the most efficient method when the target cost (i.e. the number of objective function evaluations) is about $100*N$ [$N$ being the dimension of the parameter search space, @Baeck2013].

@@

Equifinality is a common problem encountered during model calibration, where multiple parameter sets can produce equally good fits to the observed data. This issue can arise due to several factors, including limitations in the quantity or quality of available data, competing processes within the model, and parameter interactions that affect the model output in complex ways.

@@

Consequently, several parameter sets may be statistically equivalent and parameters non-identifiable. In fact, calibration repetitions gave diverging parameter values (\hyperref[fig:unfoldingplots]{Figure 4}) while being efficient in distinguishing between species presence and absence (i.e. AUC ~ 0.9, \hyperref[fig:cmaesrepAUCcal]{Figure 3}). Thus, even if the calibrated model describes the observed species distribution very well, it does not necessarily mean that parameter values are ecologically relevant. This concern is similar to the criticisms against correlative SDMs, in which parameter values and correlations that well reproduce species ranges do not necessarily describe a complex biological reality. In our case, the constraints imposed by the explicit mathematical equations embedded in the models we used were not sufficient to ensure calibration convergence towards similar solutions that would have suggested a high biological realism. However, it is worth noting that we deliberately chose large parameter ranges (although biologically realistic, i.e. corresponding to the observations made on the different processes modelled across different species) in order to give free rein to the optimization algorithm. As our goal was to assess the performance of CMA-ES objectively, we did not attempt to minimize this non-identifiability issue by restricting the parameter space. To deal with equifinality issues, an avenue to explore could be the use of multiple objective functions during model calibration to assess different aspects of the model performance. Additionally, if a closed-form likelihood can be derived, one could use a Bayesian framework to combine prior knowledge and inverse estimation of parameters to constrain the parameter space and study the nature of trade-offs between parameters [@Hartig2012; @Cailleret2020].

@@

Our goal here was to investigate the performance of CMA-ES to calibrate quite complex process-based species distribution models using species occurrence data. We did not attempt to validate our parametrizations using temporally or spatially independent data, and AUC was only used to determine if model outputs were consistent with species distributions. However, AUC is scale-invariant: it measures how well predictions are ranked rather than their absolute values. For example, with PHENOFIT, a species with a fitness of $0.8$ could be considered as absent while another one with the same fitness could be considered as present. Therefore, when it is used as an objective function for model calibration, we probably lack some precision and consistency among species’ parameters estimation. Further work could thus be conducted to examine the effects of choosing a different objective function.
